Efficient and Accurate Modeling of Flow through Sedimentary Structures with MODFLOW 6

### 2023-10-19 latex/article.tex
provost edits
Various edits, including many aimed at consistent terminology

%\title{Improving Accuracy of Simulated Flows in Steeply Dipping Layers Using full connectivity Grids with the XT3D Multi-Point Flux Approximation in MODFLOW 6}

@@

\textbf{Key words:} groundwater flow simulation, dipping model layers, full connectivity grid, multi-point flux approximation

@@

A traditional approach to representing an aquifer system with dipping hydrogeologic layers in MODFLOW is to map the hydrologic properties onto a rectilinear grid \citep{modflow84}.  \cite{hoaglund2003} refer to this as the ``grid overlay'' method.  In the grid overlay method, the boundaries of dipping layers are represented approximately, in what \cite{bardot2022} call a ``staircase'' fashion.  Another approach available in all versions of MODFLOW is to allow the top and bottom elevations of cells in the same layer of the grid to vary spatially. Following \cite{bardot2022}, in this work we call this type of grid ``vertically offset,'' and we show how to improve the accuracy of flow simulations that use vertically offset grids. The vertical offsets in top and bottom elevations between ``horizontally'' adjacent cells allows model layers to ``deform'' with the hydrostratigraphy to ``minimize the number of model layers required to simulate an aquifer system'' \citep{modflow84} relative to the grid overlay method with a rectilinear grid consisting of horizontal layers.

All versions of MODFLOW prior to MODFLOW 6 used a two-point, ``conductance-based'' flow formulation in which the flow between two adjacent cells is proportional to the difference between the heads calculated in those two cells. The conductance-based flow formulation is most accurate when the grid satisfies the ``control-volume finite-difference (CVFD) requirement'' that the straight-line connection between two cell centers must intersect the midpoint of the cell interface at a right angle \citep{narasimhan1976integrated}. However, vertically offset grids violate the CVFD requirement because lateral cell faces are vertical but cell connections are not strictly horizontal. For this reason, the MODFLOW 6 documentation \citep{modflow6gwf} warns that ``[s]teeply dipping layers generally should not be represented'' using vertical offsets. \cite{anderson2015applied} recommend a $10^{\circ}$ dip as a practical upper limit for using vertical offsets.

To improve the accuracy of flow simulations on unstructured grids, and to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D flow formulation \citep{modflow6xt3d} was introduced in MODFLOW 6. By performing flow calculations using Darcy's Law in its tensorial form based on a multi-point approximation of the head gradient vector, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid. In theory, XT3D should be able to compensate for the geometric irregularity introduced by vertical offsets between adjacent cells in the same model layer.

\cite{bardot2022} evaluated the effects of grid design and XT3D on the ability of MODFLOW 6 to accurately simulate flow through sedimentary structures by modeling an idealized, two-dimensional permeable channel or hydrogeologic layer {\color{red} (CDL: use ``permeable feature'' instead of channel?)} embedded in a nearly impermeable surrounding medium, or ``domain.'' In their plan-view benchmark, the channel lay within the horizontal plane and was angled counterclockwise relative to the southern boundary of the model domain (by $30^{\circ}$ in most cases). Constant-head boundary conditions at the ends of the channel were set to induce uniform flow along the channel in the corresponding analytical solution. The channel and surrounding domain were discretized using either a rectilinear grid, which satisfies the CVFD requirement, or a ``flexible triangular'' grid that is unstructured with triangular cells that conform to the channel boundary but does not satisfy the CVFD requirement. They found that the standard, conductance-based flow formulation gave accurate results on the rectilinear grid using isotropic hydraulic conductivity tensors only (accuracy decreased substantially in the presence of anisotropy) {\color{red} AMP: intended emphasis of ``only''? (CDL: I believe ``only'' was intentional.  I added the part about anisotropy effects in parentheses)} but was subject to significant error on the unstructured grid, whereas the XT3D flow formulation gave accurate results on both grids for both isotropic and anisotropic hydraulic conductivity tensors, as expected. However, in the transect (cross-sectional) version of their benchmark, which used a vertically offset grid to discretize the nearly impermeable domain and a permeable channel (hydrogeologic layer) inclined by $30^{\circ}$ from the horizontal, similarly poor results were observed with and without XT3D. Given the ability of XT3D to compensate for violations of the CVFD requirement, it appeared likely that another aspect of the problem formulation or the inner workings of MODFLOW 6 was responsible for the unexpected results.  As noted by \cite{bardot2022}, a preliminary analysis by two authors of the present work (Provost and Langevin) suggested that the cell connectivity offered by the vertically offset grid used for the problem was inadequate for allowing an accurate flow solution.

MODFLOW 6 supports three different types of grids: regular MODFLOW grids consisting of layers, rows, and columns (DIS),  grids discretized by vertices (DISV), and fully unstructured (DISU) grids, which are patterned after the DISU grid type in MODFLOW-USG  \citep{modflowusg}.  The DIS and DISV grids in MODFLOW 6 are ``layered'', which means that the same two-dimensional grid (in plan view) applies to all model layers.  The ``layered'' concept of DIS and DISV grids is used within MODFLOW to determine how cells are hydraulically connected.  Cells in DIS and DISV grids are automatically assigned vertical connections to overlying and underlying cells and horizontal connections to adjacent cells in the same model layer.  DISU grids are more flexible than DIS and DISV grids in that the user has complete freedom to specify the cell connectivity and the geometric properties for each connection.  DISU grids support two types of horizontal connections: a ``layered'' horizontal connection and a ``vertically staggered'' horizontal connection \citep{modflow6gwf}.  With a layered horizontal connection, the hydraulic conductance is calculated using the full saturated thickness of the two connected model cells.  Alternatively, for a staggered horizontal connection, the hydraulic conductance is calculated using only the overlap between the two cell faces. Vertically staggered horizontal connections were introduced in MODFLOW-USG \citep{modflowusg} and included in MODFLOW 6 \citep{modflow6gwf} primarily as a means to vertically subdiscretize parts of a model domain.  The grid used by \cite{bardot2022} for the steeply dipping benchmark example used layered horizontal connections.  To our knowledge, the implications of limited horizontal connectivity in layered grids for the accuracy of simulated flows in models with steeply dipping layers, and the potential for using vertically staggered grids with full cell connectivity to overcome this limitation, have not been previously appreciated or investigated.

% text removed
% For a grid that includes vertically staggered connections, which in this work we call a ``vertically staggered grid,'' some cells have nominally ``horizontal'' connections with more than one cell in some direction. Vertically staggered DISU grids allow more flexibility in assigning grid connectivity than do the more commonly used DIS and DISV grid types \citep{modflow6gwf} in MODFLOW 6.

In the remainder of this paper we first present a theoretical justification for the hypothesis that inadequate grid connectivity is primarily responsible for the inaccurate simulation of flows in the steeply dipping layer benchmark on a vertically offset grid. Based on that theory, we propose a method for improving the accuracy of the flow solution using XT3D together with vertically offset grid with full connectivity, and we evaluate the effectiveness of that approach in a set of benchmark problems similar to those of \cite{bardot2022}. Finally, we discuss the potential implications of our findings for practical groundwater models that include high permeability contrasts and steeply dipping layers.

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.6]{../figures/schem_conn_area_flux.png}
	\caption{Schematics showing (a) grid connectivity and (b) cell interface fluxes in a two-layer model of a dipping channel. Hatching denotes impermeable boundaries along the top and bottom of the channel. In (a), black circles represent cell centers. Blue lines represent hydraulic connections between cells in a layered connectivity grid. Red lines represent additional connections that must be made in order to achieve full connectivity. In (b), gray arrows represent the uniform groundwater flux the model is attempting to simulate. Blue arrows represent components of the uniform groundwater flux normal to the horizontal and vertical cell-cell interfaces in the layered connectivity grid. Red arrows represent components of the uniform groundwater flux normal to the additional vertical cell-cell interfaces introduced to represent the full connectivity.}
	\label{fig:schem-conn-area-flux}
	\end{center}
\end{figure}

Figure \ref{fig:schem-conn-area-flux}a shows a group of four cells in a two-layer, cross-sectional MODFLOW 6 model of fully saturated flow through a permeable channel (aquifer) with impermeable top and bottom boundaries. The cells have horizontal tops and bottoms and can therefore follow the dip of the channel only on average. If the grid is represented using a layered grid type (DIS or DISV) in MODFLOW 6, a cell is hydraulically connected to each adjacent cell in the same layer, regardless of whether or not the cells overlap, and with each overlying and underlying cell {\color{red} (CDL: I clarified this to indicate that cells are connected horizontally even if they do not overlap) (JM: Maybe state that this is not implicit in the formulation - that faces have to overlap. From my experience with people modelling faults...)}, as indicated by the blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In this work we call this a vertically offset grid with ``layered connectivity''.  Note that cells that have overlapping vertical faces but are in different layers (e.g., cells 1A and 2B) do not have a direct hydraulic connection in a grid with layered connectivity. The addition of such connections between cells in different layers, as indicated by the red lines that connect cell centers in figure \ref{fig:schem-conn-area-flux}a, results in a grid with ``full connectivity''.  In MODFLOW 6, layered connectivity is represented using the structured (DIS), vertex-based (DISV), or fully unstructured (DISU) grid types, whereas full connectivity can be represented only using the DISU grid type.

% CDL: it seems like we could remove this following text.  
%The ability to handle full connectivity was first introduced to MODFLOW by \cite{modflowusg} in MODFLOW-USG, and the term ``vertically staggered'' was first used by \cite{modflow6gwf} in the context of MODFLOW 6.  The term describes a grid in which which a cell can have nominally ``horizontal'' connections with multiple cells across the same cell face. In this work, ``vertically offset'' and ``vertically staggered'' are used to distinguish grids that lack nominally ``horizontal'' connections between cells in adjacent layers (red lines in Figure \ref{fig:schem-conn-area-flux}a) from those that include such connections, respectively.  In the vertically offset grid in figure \ref{fig:schem-conn-area-flux}a (blue connections only), the interfacial area across which flow occurs between adjacent cells in the same layer is based on the vertical cell thickness (the entire thickness of the channel). In the vertically staggered grid (blue and red connections), the interfacial area for nominally ``horizontal'' flow between adjacent cells is based on the area over which the two cell faces overlap. Thus, while the total area for nominally ``horizontal'' flow across a cell face is the same as for a vertically offset grid, that total area is divided between an adjacent cell in the same layer (blue connection) and an adjacent cell in the adjacent layer (red connection). In MODFLOW 6, vertically offset grids can be represented using the structured (DIS), vertex-based (DISV), or fully unstructured (DISU) grid type, whereas vertically staggered grids can be represented only using the DISU grid type.

The default method for representing the flow between two hydraulically connected model cells in MODFLOW 6 is the ``conductance-based flow formulation'' \citep{modflow6gwf}. In this formulation, which is based on the commonly used two-point flux approximation {\color{red} *** ref? ***}, the flow is proportional to the difference in the hydraulic heads computed at the two cell centers and a hydraulic conductance based on an effective hydraulic conductivity for the connection between the cells. On strictly rectilinear grids, such as structured MODFLOW grids with no vertical offset between cells in the same layer, with hydraulic conductivity that is isotropic or has anisotropy aligned with the three mutually perpendicular grid directions, the conductance-based formulation is second-order accurate \citep{dehotin2010modeling, modflow6gwf}. This is because strictly rectilinear grids satisfy the ``control-volume finite-difference (CVFD) requirement'' that the straight-line connection between two cell centers must intersect the midpoint of the cell interface at a right angle \citep{narasimhan1976integrated}. However, vertically offset grids violate the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same layer are not strictly horizontal and therefore not perpendicular to the vertical cell interfaces they intersect. For example, in {\color{red} AMP: capitalize ``Figure'' throughout} Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell interface, which compromises the accuracy of the conductance-based formulation.

The conductance-based formulation was the only flow formulation available in versions of MODFLOW prior to MODFLOW 6. As mentioned earlier, the limitations of the conductance-based formulation on vertically offset grids were well recognized. However, the ability to use unstructured grids in MODFLOW-USG and MODFLOW 6 introduced new ways to violate the CVFD requirement, and thereby render the conductance-based formulation less accurate, even in the absence of vertical offsets. The concept of a ``ghost node'' was included in MODFLOW-USG \citep{modflowusg}, and subsequently in MODFLOW 6, as a simple and optional way to improve the accuracy for grids that violated the CVFD requirement.  As an alternative to the ghost-node method for improving accuracy, and as a way to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D flow formulation \citep{modflow6xt3d} was introduced in MODFLOW 6.  XT3D formulates the flow between two cells by interpolating head values from the two cells and their neighboring cells to construct an estimate of the full, two- or three-dimensional head-gradient vector on each side of the cell interface; applying the cell conductivity tensors to obtain an estimate of the groundwater flux  normal to the interface on each side of the interface; and reconciling the two flux estimates to ensure continuity of flow across the interface. By performing flow calculations using Darcy's Law in its tensorial form, based on a ``multi-point'' approximation of the head gradient, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid.

The dipping-channel benchmark problem of \cite{bardot2022} used a vertically offset layered grid, which is the type of grid used most often in MODFLOW models. Test simulations using the conductance-based flow formulation for a steeply dipping channel (hydrogeologic layer) embedded in a domain that is six orders of magnitude less hydraulically conductive yielded significant errors in the simulated flows. The simulated groundwater flux in the middle of the channel was not along the $30^{\circ}$ incline of the channel, but nearly horizontal ($0.03^{\circ}$ incline), and its magnitude was overestimated by 15\%. Significant errors in the simulated flows were expected in this case, given that the conductance-based flow formulation does not rigorously account for violations of the CVFD requirement that the cell-cell connection be perpendicular to the cell-cell interface, and that it takes the connection length to be the horizontal distance between cell centers.  Given the ability of XT3D to account rigorously for grid connections that are not perpendicular to cell interfaces, and for the increase in connection length due to the slope of the connection, rerunning the simulation with XT3D activated could have been expected to improve the flow solution substantially. However, \cite{bardot2022} observed similarly poor results with XT3D: the flux was still nearly horizontal ($0.01^{\circ}$ incline), and its magnitude was overestimated by 12\%. Based on a preliminary analysis by two authors of the present work (Provost and Langevin), \cite{bardot2022} hypothesized that the unexpectedly large error in the flow solution with XT3D was related to inadequate grid connectivity.

The role of grid connectivity in enabling accurate simulation of flow along a dipping channel can be understood by considering Figure \ref{fig:schem-conn-area-flux}b, in which components of the groundwater flux (specific discharge) vector representative of steady, uniform flow along the channel are superimposed on each cell interface. Gray vectors represent the uniform flux oriented along the channel, which the model is attempting to simulate. Blue vectors represent the flux components normal to cell interfaces that correspond to blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered and full connectivity grids. Red vectors represent the flux components normal to cell interfaces that correspond to red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered connectivity grid.

The blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping channel, cells must exchange water not only ``horizontally'' with neighboring cells in the same layer, but also vertically with cells in the adjacent layer. Specifically, there must be flow from cells in the bottom layer to cells in the top layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the blue vectors alone are incompatible with steady, uniform flow along the channel because the exchange of water between the two model layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the bottom layer and accumulation of flow within the top layer as one moves along the channel in the direction of flow. Thus, steady, uniform flow along the channel cannot be simulated accurately without some mechanism for returning flow from the top layer to the bottom layer. The full connectivity grid provides such a mechanism by offering additional connections between cells in adjacent layers. Vertical flows from the bottom layer to the top layer, represented by vertical blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the bottom layer by nominally ``horizontal'' flows, represented by red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-layer model for simplicity, the same reasoning applies given any number of layers.

Once the full connectivity is established, it is the role of the flow formulation to account properly for the geometry of the grid, a task for which XT3D was specifically designed. However, XT3D alone cannot compensate for grid connectivity that does not provide adequate pathways for flow. This explains why using XT3D did not improve the simulation results substantially on the layered connectivity grid in the benchmark tests of \cite{bardot2022}.  In the middle section of the channel, away from the end boundary conditions, the flow solution approached steady, uniform flow the only way that it could given the limited connectivity: by suppressing vertical flows between layers, thereby rendering the overall flow approximately horizontal.

The arguments presented in this section suggest that, in addition to the use of XT3D, full connectivity can be important for obtaining an accurate flow solution in MODFLOW 6 simulations that involve steeply dipping layers. The next section presents results of simulations designed to test this hypothesis.

@@

The test problem is patterned after the cross-sectional (transect) benchmark of \cite{bardot2022}. It attempts to simulate uniform flow in an inclined permeable channel embedded in a less permeable surrounding ``domain.'' The channel is of uniform width and hydraulic conductivity and is inclined relative to the horizontal.

To simulate flow along a channel inclined at angle $\theta$ ($30^{\circ}$ for the base case), heads are specified at the centers of cells along the perimeter of the model using an analytical solution that corresponds to a unit head gradient:
\begin{equation}
\label{eqn:head_analyt_along}
h = - x \cos \theta - z \sin \theta.
\end{equation}

\noindent where $h$ is head in meters and $x$ and $z$ are the horizontal and vertical model coordinates, respectively, in meters. The hydraulic conductivity of the channel is set to 1 m/d (meter per day) so that the analytical flow solution is a groundwater flux of 1 m/d along the channel. In the base case, the conductivity of the domain is $10^{-6}$ m/d to effectively isolate the channel hydraulically.

>>>>

%\title{Improving Accuracy of Simulated Flows in Steeply Dipping Layers Using Full-Connectivity Grids with the XT3D Multi-Point Flux Approximation in MODFLOW 6}

@@

\textbf{Key words:} groundwater flow simulation, sedimentary structures, dipping layers, full-connectivity grid {\color{red} AMP: maybe ``cell connectivity''' or ``grid connectivity'' would make a better search term?}, multi-point flux approximation

@@

A traditional approach to representing an aquifer system with dipping hydrogeologic layers in MODFLOW is to map the hydrologic properties onto a rectilinear grid \citep{modflow84}.  \cite{hoaglund2003} refer to this as the ``grid overlay'' method.  In the grid overlay method, the boundaries of dipping hydrogeologic layers are represented approximately, in what \cite{bardot2022} call a ``staircase'' fashion.  Another approach available in all versions of MODFLOW is to allow the top and bottom elevations of cells in the same layer of the grid to vary spatially. Following \cite{bardot2022}, we call this type of grid ``vertically offset.'' In this work, we show how to improve the accuracy of flow simulations that use vertically offset grids. The vertical offsets in top and bottom elevations between adjacent cells within a grid layer allows each grid layer to ``deform'' with the hydrostratigraphy. This ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84} relative to the grid overlay method.

All versions of MODFLOW prior to MODFLOW 6 used a two-point, ``conductance-based'' flow formulation in which the flow between two adjacent cells is proportional to the difference between the heads calculated in those two cells. The conductance-based flow formulation is most accurate when the grid satisfies the ``control-volume finite-difference (CVFD) requirement'' that the straight-line connection between two cell centers must intersect the midpoint of the cell-cell interface at a right angle \citep{narasimhan1976integrated}. However, vertically offset grids violate the CVFD requirement because lateral cell faces are vertical but cell connections are not strictly horizontal. For this reason, the MODFLOW 6 documentation \citep{modflow6gwf} warns that ``[s]teeply dipping layers generally should not be represented'' using vertical offsets. \cite{anderson2015applied} recommend a $10^{\circ}$ dip as a practical upper limit for using vertical offsets.

To improve the accuracy of flow simulations on unstructured grids, and to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D flow formulation \citep{modflow6xt3d} was introduced in MODFLOW 6. By performing flow calculations using Darcy's Law in its tensorial form based on a multi-point approximation of the head gradient vector, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid. In theory, XT3D should be able to compensate for the geometric irregularity introduced by vertical offsets between adjacent cells in the same grid layer.

\cite{bardot2022} evaluated the effects of grid design and XT3D on the ability of MODFLOW 6 to accurately simulate flow through sedimentary structures by modeling an idealized, two-dimensional permeable feature (``channel'') in plan view and a dipping hydrogeologic layer (which \cite{bardot2022} also generically called a ``channel'') in cross-section. {\color{red} (CDL: use ``permeable feature'' instead of channel?) (AMP: Think it's ok to use channel here in reference to Kerry and Jim's plan-view model, but not after this paragraph; made changes to that effect throughout.)} In each case, the channel or hydrogeologic layer was embedded in a nearly impermeable surrounding medium, or ``domain.'' In their plan-view benchmark, the channel lay within the horizontal plane and was angled counterclockwise relative to the southern boundary of the model (by $30^{\circ}$ in most cases). Constant-head boundary conditions at the ends of the channel were set to induce uniform flow along the channel in the corresponding analytical solution. The channel and surrounding domain were discretized using either a rectilinear grid, which satisfies the CVFD requirement, or a ``flexible triangular'' grid that is unstructured with triangular cells that conform to the channel boundary but does not satisfy the CVFD requirement. They found that the standard, conductance-based flow formulation gave accurate results on the rectilinear grid using isotropic hydraulic conductivity tensors (although accuracy decreased substantially in the presence of anisotropy) but was subject to significant error on the unstructured grid. By contrast, the XT3D flow formulation gave accurate results on both grids for both isotropic and anisotropic hydraulic conductivity tensors, as expected. However, in the cross-sectional (``transect'') version of their benchmark, which used a vertically offset grid to discretize the nearly impermeable domain and a highly permeable hydrogeologic layer inclined by $30^{\circ}$ from the horizontal, similarly poor results were observed with and without XT3D. Given the ability of XT3D to compensate for violations of the CVFD requirement, it appeared likely that another aspect of the problem formulation or the inner workings of MODFLOW 6 was responsible for the unexpected results.  As noted by \cite{bardot2022}, a preliminary analysis by two authors of the present work (Provost and Langevin) suggested that the limited cell connectivity offered by the vertically offset grid was inadequate for allowing an accurate flow solution.

MODFLOW 6 supports three different types of grids: regular MODFLOW grids consisting of layers, rows, and columns (DIS);  grids discretized by vertices (DISV); and fully unstructured (DISU) grids, which are patterned after the DISU grid type in MODFLOW-USG  \citep{modflowusg}.  The DIS and DISV grids in MODFLOW 6 are ``layered'', which means that the same two-dimensional grid structure (in plan view) applies to all grid layers.  The ``layered'' concept of DIS and DISV grids is used within MODFLOW to determine how cells are hydraulically connected.  Cells in DIS and DISV grids are automatically assigned vertical connections to overlying and underlying cells and horizontal connections to adjacent cells in the same grid layer.  DISU grids are more flexible than DIS and DISV grids in that the user has complete freedom to specify the cell connectivity and the geometric properties for each connection.  DISU grids support two types of horizontal connections: a ``layered'' horizontal connection and a ``vertically staggered'' horizontal connection \citep{modflow6gwf}.  With a layered horizontal connection, the hydraulic conductance is calculated using the full saturated thicknesses of the two connected model cells.  By contrast, for a staggered horizontal connection, the hydraulic conductance is calculated using only the area of overlap between the two cell faces. Vertically staggered horizontal connections were introduced in MODFLOW-USG \citep{modflowusg} and included in MODFLOW 6 \citep{modflow6gwf} primarily as a means to vertically subdiscretize parts of a model domain.  Vertically staggered connections offer the possibility of implementing ``full'' connectivity in the sense that every cell is hydraulically connected to every other cell with which its faces overlap.

The grid used by \cite{bardot2022} for the steeply dipping benchmark example used layered horizontal connections.  To our knowledge, the implications of limited horizontal connectivity in layered grids for the accuracy of simulated flows in models with steeply dipping hydrogeologic layers, and the potential for using full cell connectivity via vertically staggered connections to overcome this limitation, have not been previously appreciated or investigated.

% text removed
% For a grid that includes vertically staggered connections, which in this work we call a ``vertically staggered grid,'' some cells have nominally ``horizontal'' connections with more than one cell in some direction. Vertically staggered DISU grids allow more flexibility in assigning cell connectivity than do the more commonly used DIS and DISV grid types \citep{modflow6gwf} in MODFLOW 6.

In the remainder of this paper we first present a theoretical justification for the hypothesis that inadequate cell connectivity is primarily responsible for the inaccurate simulation of flows in the steeply dipping hydrogeologic layer benchmark on a vertically offset grid. Based on that theory, we propose a method for improving the accuracy of the flow solution using XT3D together with vertically offset grid with full connectivity, and we evaluate the effectiveness of that approach in a set of benchmark problems similar to those of \cite{bardot2022}. Finally, we discuss the potential implications of our findings for practical groundwater models that include high permeability contrasts and steeply dipping hydrogeologic layers.

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.6]{../figures/schem_conn_area_flux.png}
	\caption{Schematics showing (a) cell connectivity and (b) cell-cell interface fluxes in a two-grid-layer model of a dipping hydrogeologic layer. Hatching denotes impermeable boundaries along the top and bottom of the hydrogeologic layer. In (a), black circles represent cell centers. Blue lines represent hydraulic connections between cells in a layered-connectivity grid. Red lines represent additional connections that must be made to achieve full connectivity. In (b), gray arrows represent the uniform groundwater flux the model is attempting to simulate. Blue arrows represent components of the uniform groundwater flux normal to the horizontal and vertical cell-cell interfaces in the layered-connectivity grid. Red arrows represent components of the uniform groundwater flux normal to the additional vertical cell-cell interfaces introduced to represent the full connectivity.}
	\label{fig:schem-conn-area-flux}
	\end{center}
\end{figure}

Figure \ref{fig:schem-conn-area-flux}a shows a group of four cells in a two-grid-layer, cross-sectional MODFLOW 6 model of fully saturated flow through a permeable hydrogeologic layer (aquifer) with impermeable top and bottom boundaries. The cells have horizontal tops and bottoms and can therefore follow the dip of the hydrogeologic layer only on average. If the grid is represented using a layered grid type (DIS or DISV) in MODFLOW 6, a cell is hydraulically connected to each adjacent cell in the same grid layer, regardless of whether or not the cells overlap, and with each overlying and underlying cell, as indicated by the blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In this work we call this a vertically offset grid with ``layered connectivity.''  Note that cells that have overlapping vertical faces but are in different grid layers (e.g., cells 1A and 2B) do not have a direct hydraulic connection in a grid with layered connectivity. The addition of such connections between cells in different grid layers, as indicated by the red lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a, results in a grid with ``full connectivity.''  In MODFLOW 6, only layered connectivity can be represented in structured (DIS) and vertex-based (DISV) grids. Fully unstructured (DISU) grids allow representation of any kind of cell connectivity, including layered or full connectivity.

The default method for representing the flow between two hydraulically connected model cells in MODFLOW 6 is the ``conductance-based flow formulation'' \citep{modflow6gwf}. In this formulation, which is based on the commonly used two-point flux approximation {\color{red} *** ref? ***}, the flow is proportional to the difference in the hydraulic heads computed at the two cell centers and a hydraulic conductance based on an effective hydraulic conductivity for the connection between the cells. On strictly rectilinear grids, such as structured MODFLOW grids with no vertical offset between cells in the same grid layer, with hydraulic conductivity that is isotropic or has anisotropy aligned with the three mutually perpendicular grid directions, the conductance-based formulation is second-order accurate \citep{dehotin2010modeling, modflow6gwf}. This is because strictly rectilinear grids satisfy the ``control-volume finite-difference (CVFD) requirement'' that the straight-line connection between two cell centers must intersect the midpoint of the cell-cell interface at a right angle \citep{narasimhan1976integrated}. However, vertically offset grids violate the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same grid layer are not strictly horizontal and therefore not perpendicular to the vertical cell-cell interfaces they intersect. For example, in Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell-cell interface, which compromises the accuracy of the conductance-based formulation.

The conductance-based formulation was the only flow formulation available in versions of MODFLOW prior to MODFLOW 6. As mentioned earlier, the limitations of the conductance-based formulation on vertically offset grids were well recognized. However, the ability to use unstructured grids in MODFLOW-USG and MODFLOW 6 introduced new ways to violate the CVFD requirement, and thereby render the conductance-based formulation less accurate, even in the absence of vertical offsets. The concept of a ``ghost node'' was included in MODFLOW-USG \citep{modflowusg}, and subsequently in MODFLOW 6, as a simple and optional way to improve the accuracy for grids that violated the CVFD requirement.  However, implementation of the ghost-node correction is highly problem dependent, and it is not obvious how to specify optimal ghost-node locations and weights on complex, unstructured grids.  As a more ``automatic'' alternative to the ghost-node method for improving accuracy, and as a way to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D flow formulation \citep{modflow6xt3d} was introduced in MODFLOW 6.  XT3D formulates the flow between two cells by interpolating head values from the two cells and their neighboring cells to construct an estimate of the full, two- or three-dimensional head-gradient vector on each side of the cell-cell interface; applying the cell conductivity tensors to obtain an estimate of the groundwater flux  normal to the interface on each side of the interface; and reconciling the two flux estimates to ensure continuity of flow across the interface. By performing flow calculations using Darcy's Law in its tensorial form, based on a ``multi-point'' approximation of the head gradient, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid.

The dipping-hydrogeologic-layer benchmark problem of \cite{bardot2022} used a vertically offset layered grid, which is the type of grid used most often in MODFLOW models. Test simulations using the conductance-based flow formulation for a steeply dipping hydrogeologic layer (which we will call the ``aquifer'') embedded in a domain that is six orders of magnitude less hydraulically conductive yielded significant errors in the simulated flows. The simulated groundwater flux in the middle of the aquifer was not along the $30^{\circ}$ incline of the aquifer as expected, but nearly horizontal ($0.03^{\circ}$ incline), and its magnitude was overestimated by 15\%. Significant errors in the simulated flows were expected in this case, given that the conductance-based flow formulation does not rigorously account for violations of the CVFD requirement that the cell-cell connection be perpendicular to the cell-cell interface, and that it takes the connection length to be the horizontal distance between cell centers.  Given the ability of XT3D to account rigorously for cell connections that are not perpendicular to cell-cell interfaces, and for the increase in connection length due to the slope of the connection, rerunning the simulation with XT3D activated could have been expected to improve the flow solution substantially. However, \cite{bardot2022} observed similarly poor results with XT3D: the flux was still nearly horizontal ($0.01^{\circ}$ incline), and its magnitude was overestimated by 12\%. Based on a preliminary analysis by two authors of the present work (Provost and Langevin), \cite{bardot2022} hypothesized that the unexpectedly large error in the flow solution with XT3D was related to inadequate cell connectivity.

The role of cell connectivity in enabling accurate simulation of flow along a dipping hydrogeologic layer can be understood by considering Figure \ref{fig:schem-conn-area-flux}b, in which components of the groundwater flux (specific discharge) vector representative of steady, uniform flow along the hydrogeologic layer are superimposed on each cell-cell interface. Gray vectors represent the uniform flux oriented along the hydrogeologic layer, which the model is attempting to simulate. Blue vectors represent the flux components normal to cell-cell interfaces that correspond to blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered-connectivity and full-connectivity grids. Red vectors represent the flux components normal to cell-cell interfaces that correspond to red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered-connectivity grid.

The blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping hydrogeologic layer, cells must exchange water not only ``horizontally'' with neighboring cells in the same grid layer, but also vertically with cells in the adjacent grid layer. Specifically, there must be flow from cells in the bottom grid layer to cells in the top grid layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the blue vectors alone are incompatible with steady, uniform flow along the hydrogeologic layer because the exchange of water between the two grid layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the bottom grid layer and accumulation of flow within the top grid layer as one moves along the hydrogeologic layer in the direction of flow. Thus, steady, uniform flow along the hydrogeologic layer cannot be simulated accurately without some mechanism for returning flow from the top grid layer to the bottom grid layer. The full-connectivity grid provides such a mechanism by offering additional connections between cells in adjacent grid layers. Vertical flows from the bottom grid layer to the top grid layer, represented by vertical blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the bottom grid layer by nominally ``horizontal'' flows, represented by red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-grid-layer model for simplicity, the same reasoning applies given any number of grid layers.

Once the full connectivity is established, it is the role of the flow formulation to account properly for the geometry of the grid, a task for which XT3D was specifically designed. However, XT3D alone cannot compensate for cell connectivity that does not provide adequate pathways for flow. This explains why using XT3D did not improve the simulation results substantially on the layered-connectivity grid in the benchmark tests of \cite{bardot2022}.  In the middle section of the hydrogeologic layer, away from the end boundary conditions, the flow solution approached steady, uniform flow the only way that it could given the limited connectivity: by suppressing vertical flows between grid layers, thereby rendering the overall flow approximately horizontal.

The arguments presented in this section suggest that, in addition to the use of XT3D, full connectivity can be important for obtaining an accurate flow solution in MODFLOW 6 simulations that involve steeply dipping hydrogeologic layers. The next section presents results of simulations designed to test this hypothesis.

@@

The test problem is patterned after the cross-sectional (transect) benchmark of \cite{bardot2022}. It attempts to simulate uniform flow in an inclined permeable hydrogeologic layer (``aquifer'') embedded in less permeable material (the ``domain''). The aquifer is of uniform width and hydraulic conductivity and is inclined relative to the horizontal.

To simulate flow along an aquifer inclined at angle $\theta$ ($30^{\circ}$ for the base case), heads are specified at the centers of cells along the perimeter of the model using a formula that corresponds to a uniform, unit head gradient:
\begin{equation}
\label{eqn:head_analyt_along}
h = - x \cos \theta - z \sin \theta.
\end{equation}

\noindent where $h$ is head in meters and $x$ and $z$ are the horizontal and vertical model coordinates, respectively, in meters. The hydraulic conductivity of the aquifer is set to 1 m/d (meter per day) so that the analytical flow solution is a groundwater flux of 1 m/d along the aquifer. In the base case, the conductivity of the domain is $10^{-6}$ m/d to effectively isolate the aquifer hydraulically.
